Static contract checking for Haskell

### 2011-10-24 paper.tex
Paper: completed the new ctrans and related it to old ctrans.
The new ctrans is nearly the same as the old, if viewed the right way, once
we fix the old trans in the ways we had already wanted.  I also added an alternate
dtrans.

Dimitrios gave an example where defining $\ctrans{f \notin c}$ as
$\lnot \ctrans{f \in c}$ is problematic: prove $\ctrans{g ~Z \in
  \cfc}$, assuming $\min{g ~Z}$, for an abstract function $g \in \cfc
\to \cfc$.  If we define 
\protect \[ 
\ctrans{g \in x:c1 \to c2} :=
\forall x. \min{g x} \to \ctrans{x \in c1} \to \ctrans{g x \in c2},
\]
then we get stuck, because we must prove $\min{Z}$ to satisfy
$\ctrans{x \in c1}$ (i.e. $\min{Z} \land \cf{Z}$; we have $\cf{Z}$ by
definition).  On the other hand, if we define
\protect \[
\ctrans{g \in x:c1 \to c2} 
:= \forall x. \min{g x} \to \lnot \ctrans{x \notin c1} \to \ctrans{g x \in c2},
\]
then we succeed, because $\min{Z}$ is irrelevant to us in satisfying
$\ctrans{x \in c1}$ (i.e. $\lnot \min{Z} \lor \cf{Z}$).

@@

Here's a start at such a translation.  I
think it's more natural to think about variance, so write
$\ctrans{}^+$ for conclusions and $\ctrans{}^-$ for assumptions.  Then
a system that works for these $\cfc$ examples is
\protect \begin{eqnarray*}
\ctrans{x \in \cfc}^+ &:=& \min{x} \to \cf{x} \\
\ctrans{x \in \cfc}^- &:=& \cf{x}\\
\ctrans{f \in x:c1 \to c2}^+ &:=& \forall x.               
  \ctrans{x \in c1}^- \to \ctrans{f x \in c2}^+\\
\ctrans{f \in x:c1 \to c2}^- &:=& \forall x. \min{f x} \to 
  \ctrans{x \in c1}^+ \to \ctrans{f x \in c2}^-
\end{eqnarray*}
Because $\Gamma,\phi \vdash \psi \equiv \Gamma \vdash \phi \to \psi$,
it's natural to think of contract assumptions in the theory as
negative, and contract goals as positive.  An example is helpful:
\protect \begin{eqnarray*}
\ctrans{g \in \cfc \to \cfc}^+ &=& \forall x. \cf{x} \to (\min{g ~x} \to \cf{g ~x}),\\
&&\text{whereas}\\
\ctrans{g \in \cfc \to \cfc}^- &=& \forall x. \min{g ~x} \to (\min{x} \to \cf{x}) \to \cf{g ~x}.
\end{eqnarray*}
Now, given $\ctrans{g \in \cfc \to \cfc}$, we may 
want to prove $\ctrans{g ~(g ~Z) \in \cfc}$ 
from $\ctrans{g \in \cfc \to \cfc}^-$ (note the negative).
So, our goal is $\min{g ~(g ~Z)} \to \cf{g ~(g ~Z)}$.
This time the proof proceeds more naturally (no EM).  
From $g$'s contract, we get $\cf{g ~(g ~Z)}$ if we can prove 
$\min{g ~Z} \to \cf{g ~Z}$.
But that follows by instantiating $g$'s contract again, and 
then using $\cf{Z}$ to get $\min{Z} \to \cf{Z}$.

Notice the assymetry between the positive and negative arrow
translations: the positive case has no $\min{f ~x}$ assumption.  I
left this $\min{}$ out because it's already in the 
conclusion of the $\cf$ example, but I certainly haven't
worked enough examples to justify it.  Moreover, this fragment
(no $e \in \{x|p\}$ yet) already fails to have a standard
property of bidirectional systems: that you can check/verify what you assume/use.
In other words, we can't prove that, forall $e$ and $c$,
\protect \[
  \ctrans{e \in c}^- \text{ implies } \ctrans{e \in c}^+.
\]
The trouble seems to come down to needing to prove $\min{e ~x} \implies
\min{x}$, which is not true a priori.  But, it's easy to add the $\min{g ~x}$ 
assumption to the positive arrow translation to recover the
 ``negative implies positive'' property.

Going back to the current translation, we consider some of it's
metatheoretic properties.
\protect \begin{enumerate}
\item $\lnot \ctrans{e \notin c}$ should behave like a double-negation
  translation, in the sense that $\ctrans{e \in c}$ should imply
  $\lnot \ctrans{e \notin c}$.  This becomes true if we modify the
  predicate case so that the implication holds (e.g. by more accurately
  modeling the operation semantics of contracts), and add a symmetric 
  $\min{f ~x}$ assumption to $\ctrans{f \notin x:c_1 \to c_2}$.  Note that
  reverse implication does not hold, e.g. $\lnot \ctrans{x \in \cfc}$ does
  not imply $\ctrans{x \in \cfc}$.
\item The ``fancy'' $\min{}$-ified translation should 
  degenerate to the ``naive'' translation when we assume $\forall x. \min{x}$.
  This is trivially true if we define the ``naive'' translatino as the ``min'' tranlation
  plus that axiom.  In symbols, we have
  \protect \[
  \Gamma^\text{min} \vdash \phi \implies \Gamma^\text{naive} \vdash \phi.
  \]  I.e., ``\min{} is sound for `naive'''.
\item We'd like to have a result  like
  \protect \[
  \text{Given some validity or consistency assumptions on $\Gamma$ and $e \in c$}, 
    \Gamma^\text{naive} \vdash \ctrans{e \in c}^\text{naive} \implies \Gamma^\text{min} \vdash \ctrans{e \in c}^\text{min}.
  \]  I.e., ``\min{} is complete for `naive', under some assumptions about 
  what we're proving''.  However, Dimitrios points out that we shouldn't expect this
  completeness theorem without some work, since we're purposefully 
  restricting the proof search to follow
  evaulation strategy, and it's far from obvious that all proofs can proceed this way.

@@

\item There may need to be some $\min{}$ assumptions in the prelude.  E.g. right now 
  it's impossble to prove $\ctrans{True \in \cfc}$, because you can't prove 
  $\min{True}$ a priori.  On the other hand, we don't need any extra 
  $\min{}$ assumption in the arrow case, since it's already assumed, at least in
  very simple examples (checked on the simple
  example that $id \in \cfc \to \cfc$).

@@

>>>>

Dimitrios gave an example where defining $\ctrans{f \notin c}$ as
$\lnot \ctrans{f \in c}$ is problematic: prove $\ctrans{g ~Z \in
  \cfc}$, assuming $\min{g ~Z}$, for an abstract function $g \in \cfc
\to \cfc$.  If we define 
\protect \[ 
\ctrans{g \in x:c1 \to c2} :=
\forall x. \min{g x} \to \ctrans{x \in c1} \to \ctrans{g x \in c2},
\]
then we get stuck, because we must prove $\min{Z}$ to satisfy
$\ctrans{x \in c1}$ (i.e. $\min{Z} \land \cf{Z}$; we have $\cf{Z}$ by
definition).  (UPDATE: Actually, this is probably just more evidence
that $\ctrans{e \in \cfc}$ is defined incorrectly: we were already
worried about that $\min{}$, because it stops us from proving
$\cf{Z}$, even when no arrows are involved).  On the other hand, if we
define the arrow case as in Figure~\ref{ctrans}, namely
\protect \[
\ctrans{g \in x:c1 \to c2} 
:= \forall x. \min{g x} \to \lnot \ctrans{x \notin c1} \to \ctrans{g x \in c2},
\]
then we succeed, because $\min{Z}$ is irrelevant to us in satisfying
$\ctrans{x \in c1}$ (i.e. $\lnot \min{Z} \lor \cf{Z}$. UPDATE: when you
notice that that implies $\min{Z} \to \cf{Z}$, (and the reverse follows from EM), the next 
paragraph might become a little dubious.  I still think the second form 
is more intuitive.).

@@

Here's a start at such a translation.  I
think it's more natural to think about variance, so write
$\ctrans{}^+$ for conclusions and $\ctrans{}^-$ for assumptions.  Then
a system that works for these $\cfc$ examples is
\protect \begin{eqnarray*}
\ctrans{x \in \cfc}^+ &:=& \min{x} \to \cf{x} \\
\ctrans{x \in \cfc}^- &:=& \cf{x}\\
\ctrans{f \in x:c1 \to c2}^+ &:=& \forall x.               
  \ctrans{x \in c1}^- \to \ctrans{f x \in c2}^+\\
\ctrans{f \in x:c1 \to c2}^- &:=& \forall x. \min{f x} \to 
  \ctrans{x \in c1}^+ \to \ctrans{f x \in c2}^-
\end{eqnarray*}
Because $\Gamma,\phi \vdash \psi \equiv \Gamma \vdash \phi \to \psi$,
it's natural to think of contract assumptions in the theory as
negative, and contract goals as positive.  An example is helpful:
\protect \begin{eqnarray*}
\ctrans{g \in \cfc \to \cfc}^+ &=& \forall x. \cf{x} \to (\min{g ~x} \to \cf{g ~x}),\\
&&\text{whereas}\\
\ctrans{g \in \cfc \to \cfc}^- &=& \forall x. \min{g ~x} \to (\min{x} \to \cf{x}) \to \cf{g ~x}.
\end{eqnarray*}
Now, given $\ctrans{g \in \cfc \to \cfc}^-$, we may 
want to prove $\ctrans{g ~(g ~Z) \in \cfc}^+$.
So, our goal is $\min{g ~(g ~Z)} \to \cf{g ~(g ~Z)}$.
This time the proof proceeds more naturally (no EM).  
From $g$'s contract, we get $\cf{g ~(g ~Z)}$ if we can prove 
$\min{g ~Z} \to \cf{g ~Z}$ (UPDATE: this is actually the same as 
before, up to using EM to rewrite the implication).
But that follows by instantiating $g$'s contract again, and 
then using $\cf{Z}$ to get $\min{Z} \to \cf{Z}$.

Notice the assymetry between the positive and negative arrow
translations: the positive case has no $\min{f ~x}$ assumption.  I
left this $\min{}$ out because it's already in the 
conclusion of the $\cf$ example, but I certainly haven't
worked enough examples to justify it.  Moreover, this fragment
(no $e \in \{x|p\}$ yet) already fails to have a standard
property of bidirectional systems: that you can check/verify what you assume/use.
In other words, we can't prove that, forall $e$ and $c$,
\protect \[
  \ctrans{e \in c}^- \text{ implies } \ctrans{e \in c}^+.
\]
The trouble seems to come down to needing to prove $\min{e ~x}$ implies 
$\min{e}$, which is not true a priori.  But, it's easy to add the $\min{g ~x}$ 
assumption to the positive arrow translation to recover the
 ``negative implies positive'' property. (UPDATE: the symmetric $\min{}$
may also be necessary when reasoning about functions that return functions: if
$f ~x = g$ then $f ~x ~y = g ~y$, but we'd need a $\min{f ~x}$ to unfold $f$'s def).

Going back to the current translation, we consider some of its
metatheoretic properties.
\protect \begin{enumerate}
\item $\lnot \ctrans{e \notin c}$ should behave like a double-negation
  translation, in the sense that $\ctrans{e \in c}$ should imply
  $\lnot \ctrans{e \notin c}$.  This becomes true if we modify the
  predicate case so that the implication holds (e.g. by more accurately
  modeling the operation semantics of contracts), and add a symmetric 
  $\min{f ~x}$ assumption to $\ctrans{f \notin x:c_1 \to c_2}$.  Note that
  reverse implication does not hold, e.g. $\lnot \ctrans{x \in \cfc}$ does
  not imply $\ctrans{x \in \cfc}$.
\item The ``fancy'' $\min{}$-ified translation should 
  degenerate to the ``naive'' translation when we assume $\forall x. \min{x}$.
  This is trivially true if we define the ``naive'' translatino as the ``min'' tranlation
  plus that axiom.  In symbols, we have
  \protect \[
  \Gamma^\text{min} \vdash \phi \implies \Gamma^\text{naive} \vdash \phi.
  \]  I.e., ``\min{} is sound for `naive'''.
\item We'd like to have a result  like: given some validity or consistency 
  assumptions on $\Gamma$ and $e \in c$,
  \protect \[
    \Gamma^\text{naive} \vdash \ctrans{e \in c}^\text{naive} \implies \Gamma^\text{min} \vdash \ctrans{e \in c}^\text{min}.
  \]  I.e., ``\min{} is complete for `naive', under some assumptions about 
  what we're proving''.  However, Dimitrios points out that we shouldn't expect this
  completeness theorem without some work, since we're purposefully 
  restricting the proof search to follow
  evaulation strategy, and it's far from obvious that all proofs can proceed this way.

@@

\item There may need to be some $\min{}$ assumptions in the prelude.  E.g. right now 
  it's impossble to prove $\ctrans{True \in \cfc}$, because you can't prove 
  $\min{True}$ a priori.  On the other hand, we don't need any extra 
  $\min{}$ assumption in the arrow case, since it's already assumed, at least in
  very simple examples (checked on the simple
  example that $id \in \cfc \to \cfc$). (UPDATE: removing the $\min{x} \land$ in $cfc$, 
  or changing it to $\min{x} \to$, is probably the solution).

@@

\note[nc]{Here's a more fleshed out alternate translation, and a
  comparison with the current translation.}
\small
\begin{verbatim}
new translation
===============

 -- you can assume both min()s when proving e in {x|p} 
[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
 -- you need to prove min(e) but then get min(p[e/x])
[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
 -- could also write this as 'e /= unr -> p[e/x] in {true,unr}'
[[e in {x|p}]]  := e = unr \/ p[e/x] in {true, unr}

[[e in cf]]+ := min(e) -> cf(e)
-- i think we can add a 'min(e) ->' here without losing anything.
[[e in cf]]- :=           cf(e)

-- should add a 'min(e x) ->' here as discussed above.
[[e in x:c1 -> c2]]+
  := forall x.             [[x in c1]]- -> [[e x in c2]]+   
[[e in x:c1 -> c2]]-
  := forall x. min(e x) -> [[x in c1]]+ -> [[e x in c2]]-

old translation (written in implication style)
==============================================

[[e in {x|p}]]       := min(p[e/x]) /\ (e = unr \/ p[e/x] = true)
not[[e notin {x|p}]] := min(p[e/x]) -> (e /= bad /\ p[e/x] /= false)

[[e in cf]]       := min(e) /\ cf(e)
not[[e notin cf]] := min(e) -> cf(e)

[[e in x:c1 -> c2]] 
  := forall x. min(e x) -> not[[x notin c1]] -> [[e x in c2]]
not[[e notin x:c1 -> c2]]
  := forall x.             [[x in c1]] -> not[[e x notin c2]]

updated cf and {x|p} rules for old translation
==============================================

[[e in {x|p}]]       := min(p[e/x]) /\ (e = unr \/ p[e/x] in {true,unr})
not[[e notin {x|p}]] := min(p[e/x]) -> (e = unr \/ p[e/x] in {true,unr})

-- what we actually discussed was adding a top level min(e) assumption, but
-- removing the min(e) goal has a similar effect.
[[e in cf]]       :=           cf(e)
not[[e notin cf]] := min(e) -> cf(e)
\end{verbatim}
\note[nc]{And what do we see?
If we identify $\in^+$ with $\lnot\notin$ and $\in^-$ with $\in$, 
then the ``new'' translation looks almost the same,
except it has more $\min{}$'s in the predicate rules.  So, have we learned
anything? Probably:
\protect \begin{enumerate}
  \item We're more confident about our suggested changes to the old translation,
        including adding a symmetric $\min{}$ assumption to the $\notin$ arrow case. 
  \item The new translation is more intuitive / we might understand how Koen came up
        with the old translation.
  \item We can answer the question about what we give as input to Equinox: for the
        new translation, it's obvious that the theory $T$ should consist of $\in^-$ axioms.
        That is consistent with giving Equinox $\in$ axioms.  
        What to give as the goal is maybe
        less obvious, since Equinox is a refutation prover, but consider: ignoring the
        strategy (refutation), we want to prove $\Gamma^{\in^-} \vdash \ctrans{e \in c}^+$.  
        But
        that's equivalent to $\Gamma^{\in^-},\lnot\ctrans{e \in c}^+ \vdash \bot$.  
        And going back
        to the old translation, the analogy is 
        $\Gamma^{\in}, \ctrans{e \notin c} \vdash \bot$.
\end{enumerate}

The predicate case for the new translation came from an example:
}
\begin{verbatim}
try:

[[e in {x|p} ]]+ = min(e) -> min(p[e/x]) -> [[e in {x|p}]]-
[[e in {x|p} ]]- = e = unr \/ p[e/x] in {true, unr}

example:

e := (\x.case x of True -> True) -- so e false = bad
c := {x|x==True} -> CF
[[e in c]]+ = forall x. min(e x) -> [[x in {x|x==True}]]- -> [[e x in CF]]+
            = forall x. min(e x) -> (x = unr \/ p in {true,unr})
                                 -> (min(e x) -> CF(e x))

min(e x) -> min(x) /\ (
         \/ (x = bad /\ (e x) = bad) 
         \/ (x = True /\ (e x) = true) 
         \/ (x = False /\ (e x) = bad)
 -- better to add the redundant x /= bad, true, false here?
         \/ (e x) = unr
         )

if x = unr then (e x) = unr and so CF(unr) and done

if x /= unr /\ p = true then we're stuck, since we can't 
unroll the definition of ==true in p

so, change the defs:

[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
[[e in {x|p}]]  := e = unr \/ p[e/x] in {true, unr}

starting again, we have min(x) from D[[]] so we get

min(p) /\ (x = unr \/ p in {true,unr})

now

 - x = unr gives (e x) = unr as before

 - for x/=unr and p in {true,unr} we unroll def (==true) 
   using min(p), for p = ((==True) x):

   min((==True) x) -> min(x) /\ ( -- e' is (==True) here
         \/ (x = bad /\ (e' x) = bad) 
         \/ (x = True /\ (e' x) = true) 
         \/ (x = False /\ (e' x) = bad)
         \/ (e' x) = unr
         )

for p = unr, we get x notin {bad,true,false}, 
and hence (e x) = unr, and we're done.

for p = true, we get x = true, 
and hence (e x) = true, and we're done.
\end{verbatim}
\note[nc]{That example also has an alternate version of the
 definition translation for functions defined by case
matching.  It's a little more concise than the old translation.
}
\begin{verbatim}
In any case, another question: 
does equinox distinguish between D[[]] like

  (x = Nil /\ f x = eNil)
  \/
  (exists y,ys. x = Cons(y,ys) /\ f x = eCons)

and D[[]] like

  (x = Nil /\ f x = eNil)
  \/
  (x = Cons(cons_1 x,cons_2 x) /\ f x = eCons[cons_1 x/y, cons_2 x/ys])

the latter is nice because we don't need to invent names (consider

  f x = case x of
          K x -> x

we can't write

  (exists x. x = K x /\ f x = x)

since the two x's equated are not the same!). 
the former is nice because it's more concise?
\end{verbatim}
